---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setoptions,echo=FALSE}
# When numbers are displayed in this document, we prefer not to use scientific
# notation like 1.5e4 unless the numbers are very large.
options(scipen=7)
```

## Loading and preprocessing the data

The data comes as a zip file containing a single comma-separated values (csv)
file.

```{r}
# Unpack the data, if not already done
dataName <- "activity.csv"
zipName <- sub(".csv", ".zip", dataName)
if (!file.exists(dataName)) {
    unzip(zipName)
}

dat <- read.csv("activity.csv",
                colClasses=c("integer", "character", "integer"))
head(dat)
```


## What is mean total number of steps taken per day?

Filter out missing values, then group by date, then sum up steps on each date.

```{r message=FALSE}
library(dplyr)
stepsByDay <- dat %>% filter(complete.cases(.)) %>% group_by(date) %>%
    summarize(steps=sum(steps))
library(ggplot2)
qplot(stepsByDay$steps, geom="histogram", xlab="Number of steps in a day")
stepsByDaySummary <- summary(stepsByDay$steps)
meanStepsByDay <- round(stepsByDaySummary["Mean"])
medianStepsByDay <- round(stepsByDaySummary["Median"])
```

The mean number of steps per day is `r meanStepsByDay`.
The median number of steps per day is `r medianStepsByDay`.


## What is the average daily activity pattern?

From the range of the "interval" variable it looks like that variable codes
hours and minutes in 24-hour format.

```{r}
range(dat$interval)
# Just the first two digits
range(floor(dat$interval/100))
# Just the last two digits
range(dat$interval %% 100)
```

So we don't want to use "interval" as is on the x-axis: there would be big gaps
between 1255 and 1300, for example, when those two intervals should be right next to
each other. Convert "interval" to a time.

```{r}
# The group_by function used below wants times in POSIXct format. The sprintf
# function here makes the number from "interval" 4 digits long (e.g. 15 becomes
# "0015") before trying to interpret it as a time of day.
dat$betterInterval <- as.POSIXct(sprintf("%04d", dat$interval),
                                 tz="GMT", "%H%M")
stepsByInterval <- dat %>% filter(complete.cases(.)) %>%
    group_by(betterInterval) %>% summarize(steps=mean(steps))

# Use the scales library to be able to format the x-axis as just times.
library(scales)
ggplot(stepsByInterval, aes(x=betterInterval, y=steps)) +
    geom_line() +
    labs(x="Interval", y="Average number of steps in interval") +
    scale_x_datetime(labels=date_format("%H:%M"))

intervalWithMaxSteps <- dat[which.max(stepsByInterval$steps),]$interval
```

Interval `r intervalWithMaxSteps` contains the maximum number of steps.


## Imputing missing values

```{r}
nrowMissingValues <- nrow(dat) - sum(complete.cases(dat))
```

There are `r nrowMissingValues` rows with missing values.

We will fill in missing values by using the mean number of steps for that
interval. We will do this in 3 steps:

1. Get the mean number of steps for each interval.
2. Create a new data set by merging the original table with the mean number of
   steps table.
3. Create a new column with no NA's by taking the value of "steps" from the
   original table, *or* if that is NA, the value from the column we added in
   step 2.

```{r}
meanSteps <- dat %>% filter(complete.cases(.)) %>% group_by(interval) %>%
    summarize(meanSteps=mean(steps))
newDat <- left_join(dat, meanSteps, by=c("interval"))
newDat$imputedSteps <- ifelse(is.na(newDat$steps), newDat$meanSteps, newDat$steps)
```

Now repeat the histogram from the first question above to check how much of a
difference imputing the data made.

```{r message=FALSE}
newStepsByDay <- newDat %>% group_by(date) %>% summarize(steps=sum(imputedSteps))
library(gridExtra)
plot1 <- qplot(stepsByDay$steps, geom="histogram", fill=I("blue"),
               main="Original (Ignore NA)",
               xlab="Number of steps in a day")
plot2 <- qplot(newStepsByDay$steps, geom="histogram", fill=I("purple"),
               main="With Imputed Values",
               xlab="Number of steps in a day")
grid.arrange(plot1, plot2, ncol=2)
```

The new plot looks like the old except for some higher bars in the middle. Were
entire days that were all NA before are being filled in? To check that:

```{r}
length(unique(stepsByDay$date))
length(unique(newStepsByDay$date))
```

The data set with imputed values has 8 days that were not there before (they
were all NA's).

```{r}
newStepsByDaySummary <- summary(newStepsByDay$steps)
newMeanStepsByDay <- round(newStepsByDaySummary["Mean"])
newMedianStepsByDay <- round(newStepsByDaySummary["Median"])
```

The new mean number of steps per day is `r newMeanStepsByDay`
(old was `r meanStepsByDay`).  
The new median number of steps per day is `r newMedianStepsByDay`.
(old was `r medianStepsByDay`).  
This is close to the original so using means by interval seems like a
workable way to impute data.


## Are there differences in activity patterns between weekdays and weekends?

```{r}
newDat$dayOfWeek <- weekdays(as.Date(newDat$date))
newDat$weekdayOrWeekend <- as.factor(ifelse(newDat$dayOfWeek=='Saturday' |
                                            newDat$dayOfWeek=='Sunday',
                                            'weekend', 'weekday'))
head(newDat)

# This plot will be much like the average daily activity pattern plot above,
# but it needs to include the weekdayOrWeekend variable in the group_by.
stepsByInterval <- newDat %>% group_by(betterInterval, weekdayOrWeekend) %>%
    summarize(imputedSteps=mean(imputedSteps))

ggplot(stepsByInterval, aes(x=betterInterval, y=imputedSteps)) +
    geom_line() +
    facet_grid(weekdayOrWeekend ~ .) +
    labs(x="Interval", y="Average number of steps") +
    scale_x_datetime(labels=date_format("%H:%M"))
```

This plot suggests that:

- People get going earlier on weekdays (around 6am compared to around 8am on
  weekends).
- Activity is more evenly spread throughout the day on weekends (the spikes are
  not as extreme).
- People stay active slightly longer on weekends (with the last spike in
  activity occuring around 8pm compared to 7pm on weekdays).
