---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setoptions,echo=FALSE}
# When numbers are displayed in this document, we prefer not to use scientific
# notation like 1.5e4 unless the numbers are very large.
options(scipen=7)
```

## Loading and preprocessing the data

The data comes as a zip file containing a single comma-separated values (csv)
file.

```{r}
# Unpack the data, if not already done
dataName <- "activity.csv"
zipName <- sub(".csv", ".zip", dataName)
if (!file.exists(dataName)) {
    unzip(zipName)
}

dat <- read.csv("activity.csv",
                colClasses=c("integer", "character", "integer"))
head(dat)
```


## What is mean total number of steps taken per day?

Filter out missing values, then group by date, then sum up steps on each date.

```{r message=FALSE}
library(dplyr)
stepsByDay <- dat %>% filter(complete.cases(.)) %>% group_by(date) %>%
    summarize(steps=sum(steps))
library(ggplot2)
qplot(stepsByDay$steps, geom="histogram", xlab="Number of steps in a day")
stepsByDaySummary <- summary(stepsByDay$steps)
```

The mean number of steps per day is `r round(stepsByDaySummary["Mean"])`.
The median number of steps per day is `r round(stepsByDaySummary["Median"])`.


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
